A Type System for Structured Aliasing in Parallel Programs

### 2012-10-30 doc/types-pldi2013/langdef.tex
More updates to the coherence section

\subsection{Serial Execution}
\label{subsec:coherenceserial}

Although it is our goal to enable parallel execution whenever possible, we briefly
consider sequential execution of a Legion program.  We show that a serial execution trivially
satisfies the interleaving criteria required by the Legion operational semantics.  Our proof
of the soundness of parallel scheduling depends on this result.
%In addition to
%sequential execution being the default behavior of the hierarchical scheduler, our proof of the
%soundness of the parallel scheduling will depend on this result.

Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
value and memory trace that result from a sequential execution of an expression are always valid executions of
Legion programs.

@@

To determine when parallel execution is safe, we start from the serial execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution and parallel execution is safe.  To
efficiently discover these cases at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
the current region mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.

@@

If whole memory traces are non-interfering, then any interleaving of them can be sorted via pairwise swaps to match the serial memory
trace and we get the result that shows parallel execution is safe:

@@

We now show that we can turn this result into an efficient runtime test that can be performed before 
the subtasks are executed.  The key to this is the privileges on Legion tasks and their ability to
provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
extended to work on privileges and sets of privileges:
$
\begin{array}{l@{}l}
\multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)\ \vee \\
& (r_1 * r_2)\ \vee \\
& (M(r_1) \cap M(r_2) = \emptyset) \\
\\
\multicolumn{2}{l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
\end{array}$

The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have exact equivalents
for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
approximated by twotests.  The first uses (logical) region disjointness constraints from the type system to 
statically infer non-interference.  The second uses the region mapping $M$ to determine
the two physical regions and performs a run-time test for disjointness.  Although still a dynamic test, it is performed
once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
this test efficiently at runtime can be found in \cite{Legion12}.
%and often need not be performed at all
%(if one of the other three conditions is shown to hold at compile-time).

@@

The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
tasks accessing aliased regions cannot run concurrently, they may execute in either order, allowing
the task issued later in program order to run as soon as its ready, even before the task issued earlier 
in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).

>>>>

\subsection{Sequential Execution}
\label{subsec:coherenceseq}

Although it is our goal to enable parallel execution whenever possible, we briefly
consider sequential execution of a Legion program.  We show that a sequential execution trivially
satisfies the interleaving criteria required by the Legion operational semantics.  Our proof
of the soundness of parallel scheduling depends on this result.
%In addition to
%sequential execution being the default behavior of the hierarchical scheduler, our proof of the
%soundness of the parallel scheduling will depend on this result.

Sequential execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
value and memory trace that result from a sequential execution of an expression are always valid executions of
Legion programs.

@@

To determine when parallel execution is safe, we start from the sequential execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution and parallel execution is safe.  To
efficiently discover these cases at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and 
the current region mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.

@@

If whole memory traces are non-interfering, then any interleaving of them can be sorted via 
pairwise swaps to match the sequential memory
trace. This gives us the following result which permits safe parallel execution:

@@

We now show that we can turn this result into an efficient runtime test for non-interface that can be 
performed task execution.  The key to this test is the privileges on Legion tasks and their ability to
provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
extended to work on privileges and sets of privileges:

$
\begin{array}{l@{}l}
\multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)\ \vee \\
& (r_1 * r_2)\ \vee \\
& (M(r_1) \cap M(r_2) = \emptyset) \\
\\
\multicolumn{2}{l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
\end{array}$

The cases where both subtasks have read-only privileges or both subtasks have reduce-only privileges have exact equivalents
for regions, which can be tested statically.  Detecting the case where the two sets of memory addresses are disjoint is 
approximated by two tests.  The first uses (logical) region disjointness constraints from the type system to 
statically infer non-interference.  The second uses the region mapping $M$ to dynamically determine the
disjointness of the two physical regions.  Although still a dynamic test, it is performed
once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
this test efficiently at runtime can be found in \cite{Legion12}.
%and often need not be performed at all
%(if one of the other three conditions is shown to hold at compile-time).

@@

The first form of relaxed coherence is the {\em atomic} mode, which specifies that although two
tasks access aliased regions cannot run concurrently, they may execute in either order, allowing
the task issued later in program order to run as soon as its ready, even before the task issued earlier 
in program order.  This relaxation only applies if all aliased regions are annotated with atomic coherence.
%or a more relaxed coherence mode (see Section~\ref{subsec:coherencesimult}).
